The yaq Project: \\ A Standard for Scientific Instrumentation

### 2022-10-26 paper.tex

\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}

@@

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing hardware availability has brought greater instrument complexity to small research groups.
The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision.
Highly modular hardware components enable extreme experimental flexibility.
Herein we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported with tooling available to create more.
The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components, purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The process of integrating these is an important but time-consuming step of modern instrument and systems development.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

@@

We have created a novel network-based standard for scientific instrumentation, called \yaq{}.  This standard borrows the most important ideas from
 established projects used by large user facilities while retaining simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.

Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
Next, we discuss how inflexible orchestration software can limit experimental creativity.
Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
Finally, we discuss the heavy software maintenance burden that many instrument designers face.
In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
This paper provides an overview of the concepts and motivations behind \yaq{}.
Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected two two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

@@

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.

The second hardware interface barrier: Reduction of the interfacing code to an easier-to-manage subset of that available.  Here, we implement an approach useful for
systems integration in the sciences which we call a Minimal Programming Interface (MPI).   The MPI is essentially a reduced set of functions found in the hardware API.   The MPI 
 is defined as the minimal set of API functions for a given hardware component concluded by the researchers as necessary for the fuller system development. 
 The value of this MPI is that it can be formed by wrapping and/or converting the API from its native language into a language known to be transparent to researchers, such as the programming
 language Python.  The \yaq{} traits feature acts as a tool in exposing the MPI from the base APIs often supplied by the hardware manufacturer, as the traits
 automatically define the primary set of key functions, and any additionally required functions are added by the researcher on top of the primary set.

The third hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
For example, the Windows XP computer which has the obsolete interface card and drivers can be placed into a private network with only \yaq{} interface communication.
Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
A client running up to date Python 3.10 communicates seamlessly with a daemon running Python 3.7.

The fourth hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figures/lines_histogram},
	\caption{  \label{fig:histogram} Histogram of the number of lines for each implemented daemon.
	Some daemons are implemented in ways that share code, resulting in apparent line counts less than 10.
	For example, the Thorlabs APT\cite{thorlabs_apt} motor implementation supports at least eight different daemons with each specifying only a handful of constants.
	These are extreme examples which are not representative of most hardware interfaces, so we omit them here.
	}
\end{figure}

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}

>>>>

\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1} % consider using RevTex4.2

@@

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing modular, commercial hardware availability has brought greater instrument complexity to small research groups.
The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision even when highly modular hardware components enable extreme experimental flexibility.
Herein, we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex in
\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported with tooling available to create more.
The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components that are both purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

@@

We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.

Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
Next, we discuss how inflexible orchestration software can limit experimental creativity.
Then, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
Finally, we discuss the heavy software maintenance burden that many instrument designers face.
In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
This paper provides an overview of the concepts and motivations behind \yaq{}.
Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected to two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

@@

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system. % consider making this sentence (and the next paragraph starts) subsection headings 
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.

The second hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
For example, the Windows XP computer which has the obsolete interface card and drivers can be placed into a private network with only \yaq{} interface communication.
Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
A client running up to date Python 3.10 communicates seamlessly with a daemon running Python 3.7.

The third hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figures/lines_histogram},
	\caption{  \label{fig:histogram} Histogram of the number of lines for each implemented daemon.
	Some daemons are implemented in ways that share code, resulting in apparent line counts less than 10.
	For example, the Thorlabs APT\cite{thorlabs_apt} motor implementation supports at least eight different daemons with each specifying only a handful of constants.
	These are extreme examples which are not representative of most hardware interfaces, so we omit them here. % it's not clear to me why these instances are not included (and scaled down???) It is a _feature_ not a bug that one can implement motor support with tiny amounts of code.
	}
\end{figure}

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.

While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in fewer than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}
